---
title: "Summary Graphs of NUTR630 Intake"
author: "Dave Bridges, Liv Anderson and Noura El Habbal"
date: "Septembe 3, 2018"
output:
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: no
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```


```{r data-input}
library(readr)
filename <- 'https://docs.google.com/spreadsheets/d/e/2PACX-1vSDRxu3Ur53iZVsg5Gg9nArNiKY2-xguRzoeWl-wQ5Ky8qxRGnuXXMWSReSF1tR_SU6dmcBTgb-cQiY/pub?gid=830256665&single=true&output=csv'
data <- read_csv(filename)
```

These data can be found in `r getwd()` in a file named `r filename`.  This script was most recently updated on `r date()`.

# Analysis

## What Majors

```{r majors-summary}
library(forcats)
#grouped with most common 4

count.majors <-
  data %>%
  mutate(`Which discipline most closely matches your undergraduate degree?` = fct_lump(as.factor(data$`Which discipline most closely matches your undergraduate degree?`), n=4)) %>%
  mutate(`Which discipline most closely matches your undergraduate degree?` = 
           fct_recode(`Which discipline most closely matches your undergraduate degree?`, 
                      "Neuroscience" = "Psychology-related (Psychology, Neuroscience, etc.)",
                      "Nutrition" = "Food Quality & Safety",
                      "Nutrition" = "Nutrition & dietetics (minor in biology)",
                      "Nutrition" = "Dietetics",
                      "Nutrition" = "Nutrition and Food Science",
                      "Nutrition" = "Food & Nutritional Sciences",
                      "Biological Sciences" = "Biochemistry",
                      "Humanities" = "English Lit and Commuications, years later I took biochem pre-reqs for my program")) %>%
  group_by(`Which discipline most closely matches your undergraduate degree?`) %>%
  count() %>%
  arrange(desc(n)) 
  
with(count.majors, barplot(n,
                          las=1,
                          cex.names=0.66,
                          main="What was your major?",
                          col=color.scheme,
                          names.arg=`Which discipline most closely matches your undergraduate degree?`))

library(ggplot2)
ggplot(count.majors,aes(y=n,x=reorder(`Which discipline most closely matches your undergraduate degree?`,-n))) +
         geom_bar(stat='identity',fill=color.scheme[1]) +
  labs(y="Number of Students",
       title="Which discipline most closely matches your undergraduate degree?",
       x="")
```
  
  
## What Topics are Students Interested In?

```{r student-interests}
library(sjPlot)

student.interest.data <- 
  data %>%
  select(starts_with('Please answer these questions about your interests')) %>%
  rename("Biochemistry - Interest" = "Please answer these questions about your interests [Macronutrient biochemistry is of interest to me.]",
         "Biochemistry - Important" = "Please answer these questions about your interests [Macronutrient biochemistry is important for my career interests.]",
         "Digestion - Interest" = "Please answer these questions about your interests [Comprehensive understanding of the digestive tract is of interest to me.]",
         "Digestion - Important" = "Please answer these questions about your interests [Comprehensive understanding of the digestive tract is important for my career interests.]") %>%
  mutate(`Biochemistry - Interest`= fct_recode(`Biochemistry - Interest`, 
                                                 '1'="Strongly Agree",
                                                 '2'="Agree",
                                                 '3'="Neutral",
                                                 '4'="Disagree",
                                                 '5'="Strongly Disagree")) %>%
  mutate(`Biochemistry - Important`= fct_recode(`Biochemistry - Important`, 
                                                 '1'="Strongly Agree",
                                                 '2'="Agree",
                                                 '3'="Neutral",
                                                 '4'="Disagree",
                                                 '5'="Strongly Disagree")) %>%
    mutate(`Digestion - Interest`= fct_recode(`Digestion - Interest`, 
                                                 '1'="Strongly Agree",
                                                 '2'="Agree",
                                                 '3'="Neutral",
                                                 '4'="Disagree",
                                                 '5'="Strongly Disagree")) %>%
      mutate(`Digestion - Important`= fct_recode(`Digestion - Important`, 
                                                 '1'="Strongly Agree",
                                                 '2'="Agree",
                                                 '3'="Neutral",
                                                 '4'="Disagree",
                                                 '5'="Strongly Disagree")) %>%
  mutate(`Biochemistry - Interest`=as.numeric(as.character(`Biochemistry - Interest`))) %>%
  mutate(`Biochemistry - Important`=as.numeric(as.character(`Biochemistry - Important`))) %>%
  mutate(`Digestion - Interest`=as.numeric(as.character(`Digestion - Interest`))) %>%
  mutate(`Digestion - Important`=as.numeric(as.character(`Digestion - Important`))) 

plot_likert(student.interest.data,
           sort.frq=NULL,
           values='hide',
           reverse.colors=TRUE,
           show.legend=FALSE,
           show.n=FALSE)
```

### What Topic are you Most Interested In?

```{r qualitative-interests}
library(stringr)
topic.interest <- 
  data %>%
  select(`After reading the syllabus, which topic are you most excited to learn about?`) %>%
  mutate(Interest = `After reading the syllabus, which topic are you most excited to learn about?`) %>%
  mutate(Microbiome = ifelse(str_detect(Interest, "Microbiome|microbiome"), TRUE, FALSE),
         Fiber = ifelse(str_detect(Interest, "Fiber|fibre"), TRUE, FALSE),
         Balance = ifelse(str_detect(Interest, "balance|Balance"), TRUE, FALSE),
         Digestion = ifelse(str_detect(Interest, "digest|Digest"), TRUE, FALSE),
         Biochem = ifelse(str_detect(Interest, "biochem|Biochem"), TRUE, FALSE)) %>%
  select(Microbiome,Fiber,Balance,Digestion,Biochem) %>%
  gather(key=Interest, value=Response) %>%
  filter(Response==TRUE) %>%
  count(Interest)

ggplot(topic.interest, aes(y=n,x=reorder(Interest,n))) +
  geom_bar(stat='identity',fill=color.scheme[1]) +
  coord_flip() +
  labs(y='Number of Students',
       x="",
       title="")
```

## Applications of Course Content

```{r applications-content}
library(stringr)
application.data <- 
  data %>%
  select(`What applications of this course content are you interested in?  Select all that apply.`) %>%
  rename(Applications = `What applications of this course content are you interested in?  Select all that apply.`)

application.results = str_split(application.data$Applications, pattern=', ', simplify =F) %>% unlist() 

application.summary <-
  as.data.frame(application.results) %>%
  rename(Results = application.results) %>%
  count(Results) %>%
  arrange(n)

ggplot(application.summary, aes(y=n,x=reorder(Results,n))) +
  geom_bar(stat='identity',fill=color.scheme[1]) +
  coord_flip() +
  labs(y='Number of Students',
       x="",
       title="")
```

# Student Experience Data

```{r student-experience}
student.exp.data <-
  data %>%
  select(`I have previously written a literature review for a class.`,
`In college, have you ever done an in-class presentation?`,`Have you ever been asked to review a peer's presentation?`,`Have you ever been asked to evaluate a news article for class?`,`Have you ever written review questions based on material you were taught?`)

countYes = function(v){length(v[v=="Yes"])}
student.exp.data.summary <- sapply(student.exp.data,countYes) 
student.exp.data.summary <- as_tibble(student.exp.data.summary)
student.exp.data.summary$Question <- rownames(student.exp.data.summary) 

ggplot(student.exp.data.summary, aes(x=reorder(Question,value),y=value/dim(student.exp.data)[1]*100)) +
  geom_bar(stat='identity',fill=color.scheme[1]) + coord_flip() +
  labs(x="",y="Percent",title="Assignment Experiences")
```

## Online Resources

```{r online-resources}
online.data <-
  data %>%
  select(starts_with("Which of the following online resources do you use for learning about nutrition?  Have you ever shared nutrition content in them?"))
colnames(online.data) <- gsub(".*\\[|\\]", "", colnames(online.data))

online.data.summary <-
  gather(online.data,key="Tool",value="Use") %>% 
  group_by(Tool) %>% 
  count(Use) %>%
  na.omit 
  
ggplot(filter(online.data.summary, Use != "Contributor"), aes(y=n,x=reorder(Tool,n))) +
  geom_bar(stat='identity',fill=color.scheme[1]) +
  facet_grid(Use~.) +
  labs(y="Number of Students",x="",title="Use of Online Resources")
```

## GradeCraft Familiarity

```{r gradecraft}
gradecraft <- 
  data %>%
  group_by(`Are you familiar with GradeCraft?`) %>%
  count()

with(gradecraft, barplot(n,
                          las=1,
                          ylab="Responses",
                          main="Are you familiar with GradeCraft?",
                          col=color.scheme,
                          names.arg=`Are you familiar with GradeCraft?`))
```

Only `r gradecraft$n[2]` out of `r gradecraft$n[2]+gradecraft$n[1]` students were familiar with GradeCraft.

# Session Information

```{r session-information, echo=T}
sessionInfo()
```