---
title: "Saving Data and Plots - R Gapminder"
author: "Dave Bridges"
date: "4/23/2018"
output:
  html_document:
    keep_md: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r sample-plot}
library(ggplot2)
library(gapminder)
ggplot(data = gapminder, aes(x = gdpPercap, fill=continent)) +
 geom_density(alpha=0.6) + facet_wrap( ~ year) + scale_x_log10()
```

You can save a plot from within RStudio using the ‘Export’ button in the ‘Plot’ window. **Show this** This will give you the option of saving as a .pdf or as .png, .jpg or other image formats.

```{r saving-plots}
pdf("Life_Exp_vs_time.pdf", width=12, height=4)
ggplot(data = gapminder, aes(x = gdpPercap, fill=continent)) +
 geom_density(alpha=0.6) + facet_wrap( ~ year) + scale_x_log10()
dev.off()
```

This can be repeated with other commands including png or jpg.  **You must remember to close the device**  

 Within ggplot there is another method

# Saving within ggplot

```{r ggsave}
ggplot(data = gapminder, aes(x = gdpPercap, fill=continent)) +
 geom_density(alpha=0.6) + facet_wrap( ~ year) + scale_x_log10()

ggsave("My_most_recent_plot.pdf")
```

# What about saving data

can write as table, excel or csv

```{r saving-data}
library(gapminder)
gapminder

write.csv(gapminder, file="Gapminder_data.csv")
```