---
title: "Saving Data and Plots - R Gapminder"
author: "Dave Bridges"
date: "4/23/2018"
output:
  html_document:
    keep_md: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r sample-plot}
library(ggplot2)
library(gapminder)
ggplot(data = gapminder, aes(x = gdpPercap, fill=continent)) +
 geom_density(alpha=0.6) + facet_wrap( ~ year) + scale_x_log10()
```

# Several Ways to Save Plots (2:00-2:05)

You can save a plot from within RStudio using:

* the `pdf()`, `png()`, `jpg()`
* the ‘Export’ button in the ‘Plot’ window. **Show this** This will give you the option of saving as a .pdf or as .png, .jpg or other image formats.
* `ggsave()` see `?ggsave()`
* within a Rmd file (will talk about later)

```{r saving-plots}
pdf("Life_Exp_vs_time.pdf", width=12, height=4)
ggplot(data = gapminder, aes(x = gdpPercap, fill=continent)) +
 geom_density(alpha=0.6) + facet_wrap( ~ year) + scale_x_log10()
dev.off()
```

This can be repeated with other commands including png or jpg.  **You must remember to close the device**  

 Within ggplot there is another method

# Saving within ggplot

```{r ggsave}
ggplot(data = gapminder, aes(x = gdpPercap, fill=continent)) +
 geom_density(alpha=0.6) + facet_wrap( ~ year) + scale_x_log10()

ggsave("My_most_recent_plot.pdf")
```

# What about saving data (2:05-2:15)

can write as table, excel or csv

```{r saving-data}
library(gapminder)
gapminder

write.csv(gapminder, file="Gapminder_data.csv")
```

Can also write as a table (tab separated or excel file).  To use excel I prefer the packages `readxl` and `writexl` which contain the functions `read_excel` and `write_xlsx`.

# Challenge (2:15-2:20)

```{saving-challenge}
Write a data-cleaning script file that subsets the gapminder data to include only data points collected since 1990.

Use this script to write out the new subset to a file in the cleaned-data/ directory.
```

# Up Next: Break and dplyr