---
title: "Informative title"
author: "Authors"
date: "Date script was first written"
output:
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: yes
    toc: yes
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=FALSE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```

# Purpose

Graphs for NUTR630

## Effects of LCHF RCT on Cholesterol Levels

```{r ldl-c-rct}
ldl.data.foster <- data.frame(Treatment=c("LCHF","HCLF"), LDL=c(5.4,-7.4), LDL.CI=c(19.2,16.6)) %>% #from Foster NEJM 2003
  mutate(LDL.SE = LDL.CI/1.96)

library(ggplot2)
ggplot(ldl.data.foster, aes(y=LDL,
                     x=Treatment,
                     ymin=LDL-LDL.CI,
                     ymax=LDL+LDL.CI)) +
  geom_bar(stat='identity') +
  geom_errorbar(width=0.5) +
  geom_hline(yintercept=0, lty=2) +
  labs(y="Change in LDL-Cholesterol",
       title="Effects of High or Low Carbohydrate Diet on LDL-Cholesterol",
       subtitle="From Foster et al JAMA 2018")


data.gardner <- data.frame(Treatment=c("LCHF","HCLF"), 
                               LDL=c(3.62,-2.12), 
                               LDL.CI=c(6.19-3.62,4.7-2.12),
                               HDL=c(2.64,0.40),
                               HDL.CI=c(1.18-0.4,3.41-2.64)) %>% #from Gardner et al JAMA 2018
  mutate(LDL.SE = LDL.CI/1.96,
         HDL.SE=HDL.CI/1.96,
         TC=HDL+LDL,
         TC.SE=sqrt(LDL.SE^2+HDL.SE^2))

ggplot(data.gardner, aes(y=LDL,
                     x=Treatment,
                     ymin=LDL-LDL.CI,
                     ymax=LDL+LDL.CI)) +
  geom_bar(stat='identity') +
  geom_errorbar(width=0.5) +
  geom_hline(yintercept=0, lty=2) +
  labs(y="Change in LDL-Cholesterol",
       title="Effects of High or Low Carbohydrate Diet on LDL-Cholesterol",
       subtitle="From Gardner et al JAMA 2018")

ggplot(data.gardner, aes(y=TC,
                     x=Treatment,
                     ymin=TC-TC.SE,
                     ymax=TC+TC.SE)) +
  geom_bar(stat='identity') +
  geom_errorbar(width=0.5) +
  geom_hline(yintercept=0, lty=2) +
  labs(y="Change in Blood Cholesterol",
       title="Effects of High or Low Carbohydrate Diet on Cholesterol",
       subtitle="From Gardner et al JAMA 2018")
```

This script was most recently updated on **`r date()`**.

# Analysis

Describe the analysis as you intersperse code chunks

# Interpretation

A brief summary of what the interpretation of these results were

# Session Information

```{r session-information, echo=T}
sessionInfo()
```

# References

If needed, using Rmarkdown citation tools (see this link for more information: http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html)