---
title: "M-Sci Notes Lecture 1"
author: "Dave Bridges"
date: "6/27/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Who Am I?

* Professor in the department of Nutritional Sciences at the University of Michigan School of Public Health
* Born in Canada, did my undergraduate degree in Biochemistry, was interested in medicine, then in science

## Why are analytical skills important

* When I was an undergrad in the late 90s my University education was about facts and the professors and textbooks were the gateway to those facts.  Wikipedia, Youtube, Reddit did not exist.  As an example, memorizing glycolysis or the structures of amino acids.
* Now my students have access to **tons** of information, and tons of facts.  We as teachers are striving to provide you with skills to organize those facts.
* Analytical skills are essential to finding your **own** data, and asking your **own** questions.  Learning can be much more active this way.

## How do I use R in my research

* We study the mechanisms of chronic disease, things like obesity, diabetes and cancer.  We try to disect the molecular pathways that cause the diseases and understand how we can modify these.
* A lot of our research involves analyzing public datasets.  These could be genetic data, or metabolite data with thousands of compounds, or how tens of thousands of transcripts might change after a drug treatment.  A lot of our research is in making sense out of these numbers.
* We do all of our data analysis, do all of our statistical analyses and make all of our graphs in R.  We then share all the raw data, along with how we made these graphs openly with other researchers.

Lesson is from http://www.datacarpentry.org/R-ecology-lesson/index.html

# Before We Start

Follows the plan in http://www.datacarpentry.org/R-ecology-lesson/00-before-we-start.html

Need to cover these learning objectives:

* Describe the purpose of the RStudio Script, Console, Environment, and Plots panes.
* Organize files and directories for a set of analyses as an R Project, and understand the purpose of the working directory.

## What is R and What is RStudio

* R is both a programming language and software that interprets it.
* RStudio is an interface to using R.  
* For what we are doing you need to have both R and RStudio installed on your computer
* You may be more familiar with pointing and clicking software like excel, R is text based.  That makes the learning curve for R much steeper, but the end result is that you are much more powerful.  If you are doing a complex procedure, it is hard to document things like click here, then enter this, then click there.  *Scripts* are much more reproducible and make your procedures very clear.  **Does anyone have a good definition for what a script might be?**
* Scripts are also shareable, which means that if it runs on your computer it should run on my computer, and if you have a mistake, we can share the text and fix it.  You will see that over the next few days.  This makes scripts very *reproducible*, **can you think of any good reasons why reproducibility should be important in the sciences**
* R can work on all types of data, maybe you have just a few lines of code, or maybe you are interrogating a dataset with several million rows.  R works the same on both.
* How can you get help?  There are many places online such as stack overflow, mailing lists, or local meetup groups that people use to learn new skills.  R has a very open and welcoming community.  
* `Write on the board`: R is free, open source, and cross platform.  **What do you think these mean, and why are they relevant?**

## Navigating RStudio

* Open up R Studio, you should see 4 panes, clockwise from top left.  **Does anyone not see four panes?**
* **Source**.  This is where you will write and save your scripts.
* **Environment and History**.  This is information about the data you are working with
* **Help and Viewer**.  Shows graphs and the very important help panes.
* **Console**.  If you open R without RStudio this is the only part you have.  This is where code is executed.

## Getting Set Up and the Working Directory

* Under File click New Project, then Click New Directory, then New Project.  Call it **R for M-Sci** and make it a subdirectory of your Desktop.  This is where we are going to keep all your scripts, data and graphs for the next few days.
* The working directory is also a way to organize your data, you might want some folders in your working directory. We want to creat directories called data, data_output, documents, figure_output, scripts **Write this on the board ad an image**.  You could do this in windows explorer or on the terminal but we can also do this within R.  In the bottom left console enter these commands just as I type them.

```{r create-directories, warning=FALSE}
dir.create("data")
dir.create("data_output")
dir.create("documents")
dir.create("figure_output")
dir.create("scripts")
```

Look in the Files pane on the bottom right, you should see a few new directories.  Congratulations you just executed your first R code.  Lets look at what you actually did.  You called a *function* dir.create and you gave it an argument (the directory name).  The function then did something to create that folder.  

* Lets type dir.create and run that without an argument.  
* This tells us what the function is.  
* There are several arguments, some of which already have default values.  
* Dont worry about what these mean just yet, but you can see there are arguments for path, showWarnings, recursive and mode

But how do i know what I just did?  

* Look at the History tab it shows all the commands that you entered.  
* This can be helpful, but you are not tracking your code in a script, which would be better.  
* Instead lets enter this R code in the top left panel where we keep scripts.
* To run a script you can click on the line and hit CONTROL-ENTER, **do you see anything**.  This is a warning, everything is ok its just reminding you as a helpful note.  **If anyone has anything else, or did not see the directories in the file pane please raise your hand**
* You can also select multiple lines and run them.  
* Note that what this did was move the lines from the top left scripts panel into the bottom left console panel.  We always want to work in the top left panel.
* One last point, see how there is a `>` on the command line?  That means the console is ready to recieve a command.  If you see a `+` sign that means that it expects something else.  Often this is a missing quotation mark.  Very common mistake, I do this all the time!  To fix it, press ESCAPE
 and try again.

**IF AHEAD TRY LOOKING AT THE HELP CONSOLE.**  Should be 10:30 at this point.

Note from Meg

<quote>While going through this, remind the students about the “your primary collaborator is yourself 6 months from now, and your past self doesn’t answer email” axiom. Note that this is part of why using R makes sense (since having a script makes it much easier to figure out what they did in the past!),</quote> 

# Introduction to R

Once you’ve covered those two objectives, move on to the “Introduction to R” portion of the lesson.  This can be found at http://www.datacarpentry.org/R-ecology-lesson/01-intro-to-r.html. For this first day, focus on just the “Creating objects in R” section and associated learning objectives:

* Define the following terms as they relate to R: object, assign, call, function, arguments, options.
* Assign values to objects in R.
* Learn how to name objects
* Use comments to inform script.
* Solve simple arithmetic operations in R.
* Call functions and use arguments to change their default options.

## R as a calculator

R can be used as a calculator, you can do several types of arithmetic.  Notice that when i run these, the output is the result

```{r calculator}
3+5 # add
123-32 # subtract
7*8 # multiply
132/11 # divide
3^4 # power
12 == 3*4 # conditional
12 != 3*4 # anticonditional
12 >= 3*4 #greater than or equal to
'my jokes' == 'funny' #can be done with text too
```

# Closing thoughts

Remind students about homework due before next class

