---
title: "Analysis of Teaching Evaluations"
author: "Dave Bridges"
date: "1/28/2018"
output:
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)

opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```

## Data Import

Copied evaluation table, added quotes around the questions and imported into excel.  Converted this to a csv file for import.  Did this for both 2016 and 2017 teaching evaluations, downloaded from wolverine access.

```{r data-import}
library(readr)
library(dplyr)
data.2016.datafile <- '2016 Evaluations.csv'
data.2017.datafile <- '2017 Evaluations.csv'

input_col_types <- cols(
  Number = col_factor(levels=NULL),
  Question = col_factor(levels=NULL))

data.2016 <- read_csv(data.2016.datafile, col_types=input_col_types) %>% mutate(Year="2016")
data.2017 <- read_csv(data.2017.datafile, col_types=input_col_types)  %>% mutate(Year="2017")

te.data.wide <- full_join(data.2016,data.2017, 
                     by = c("Number", "Question"), 
                     suffix = c(".16", ".17"))

te.data <- 
  rbind(data.2016,data.2017) %>%
  mutate(Total = SD+D+N+A+SA) %>%
  mutate(`Strongly Disagree`=SD/Total*100,
         `Disagree`=D/Total*100,
         `Neutral`=N/Total*100,
         `Agree`=A/Total*100,
         `Strongly Agree`=SA/Total*100)
```

The imported datafiles include:

* `r data.2016.datafile`
* `r data.2017.datafile`

# Overall Questions

## Overall, this was an excellent course.

```{r excellent-course}

overall.data <-
  te.data %>%
  filter(Number==1) %>%
  mutate(Item=as.factor(Year))

library(tidyr)
library(dplyr)

individualized.overall.data.agg <-
  te.data %>%
  dplyr::select(Question, Year, SA, A, N, D, SD) %>%
  dplyr::select(SA:SD,Question,Year) %>%
  gather(value=Number,key=Response,-Question,-Year) %>%
  group_by(Question,Year,Response) %>%
  expand(Count=seq(1:Number)) %>%
  mutate(Value = ifelse(Response=='SA', 5,
                        ifelse(Response=='A',4,
                               ifelse(Response=='N',3,
                                      ifelse(Response=='D',2,
                                             ifelse(Response=='SD',1,0))))))
  

question.summary <-
  individualized.overall.data.agg %>%
  group_by(Question,Year) %>%
  summarize(value.list = list(Value)) %>%
  spread(Year,value.list) %>%
  rename(First.Year=`2016`,
         Second.Year=`2017`) %>%
  group_by(Question) %>%
  mutate(First.mean=mean(unlist(First.Year)),
         Second.mean=mean(unlist(Second.Year))) %>%
  mutate(Change=Second.mean-First.mean) %>%
  filter(First.Year != 'NULL') %>%
  mutate(Mann.Whitney.p=wilcox.test(unlist(First.Year),unlist(Second.Year))$p.value,
         Mann.Whitney.t=wilcox.test(unlist(First.Year),unlist(Second.Year))$statistic) %>%
  arrange(Mann.Whitney.p) %>%
  dplyr::select(-First.Year,-Second.Year)

kable(question.summary,caption="Mann-Whitney Tests for teaching evaluation questions asked in both years.")
write.csv(question.summary, file="Statistical Tests for Teaching Evaluations.csv")

library(HH)

plot.data <- overall.data[7:3]
rownames(plot.data) <- c("Before","After")

likert(plot.data, horizontal = FALSE,
       main = "Overall, this was an excellent course.",
       xlab = "Percent", # becomes ylab due to horizontal arg,
       ylab = c("Before","After"),
       title = "Overall, this was an excellent course.",
       auto.key = list(space = "right", columns = 1,
                     reverse = TRUE))
```

# Learning

## My interest in the subject has increased because of this course.

```{r increased-interest}
overall.data <-
  te.data %>%
  filter(Number==1632) %>%
  mutate(Item=as.factor(Year))

plot.data <- overall.data[7:3]
rownames(plot.data) <- c("Before","After")

likert(plot.data, horizontal = FALSE,
       main = "My interest in the subject has increased because of this course.",
       xlab = "Percent", # becomes ylab due to horizontal arg,
       ylab = c("Before","After"),
       title = "Overall, this was an excellent course.",
       auto.key = list(space = "right", columns = 1,
                     reverse = TRUE))
```

## I learned a great deal from this course.

```{r learned-a-great-deal}
overall.data <-
  te.data %>%
  filter(Number==3) %>%
  mutate(Item=as.factor(Year))

plot.data <- overall.data[7:3]
rownames(plot.data) <- c("Before","After")

likert(plot.data, horizontal = FALSE,
       main = "I learned a great deal from this course..",
       xlab = "Percent", # becomes ylab due to horizontal arg,
       ylab = c("Before","After"),
       title = "Overall, this was an excellent course.",
       auto.key = list(space = "right", columns = 1,
                     reverse = TRUE))
```


## This course advanced my understanding of the subject matter.

```{r advanced-understanding}

overall.data <-
  te.data %>%
  filter(Number==1631) %>%
  mutate(Item=as.factor(Year))

plot.data <- overall.data[7:3]
rownames(plot.data) <- c("Before","After")

likert(plot.data, horizontal = FALSE,
       main = "This course advanced my understanding of the subject matter.",
       xlab = "Percent", # becomes ylab due to horizontal arg,
       ylab = c("Before","After"),
       title = "Overall, this was an excellent course.",
       auto.key = list(space = "right", columns = 1,
                     reverse = TRUE))
```

## I learned a great deal from this course.

```{r learned-great-deal}

overall.data <-
  te.data %>%
  filter(Number==3) %>%
  mutate(Item=as.factor(Year))

plot.data <- overall.data[7:3]
rownames(plot.data) <- c("Before","After")

likert(plot.data, horizontal = FALSE,
       main = "I learned a great deal from this course.",
       xlab = "Percent", # becomes ylab due to horizontal arg,
       ylab = c("Before","After"),
       auto.key = list(space = "right", columns = 1,
                     reverse = TRUE))
```

# Fairness in Grading

## The grades in this course were fairly determined.

```{r grades-fairly-determined}
overall.data <-
  te.data %>%
  filter(Number==894) %>%
  mutate(Item=as.factor(Year))

plot.data <- overall.data[7:3]
rownames(plot.data) <- c("Before","After")

likert(plot.data, horizontal = FALSE,
       main = "The grades in this course were fairly determined.", # or give "title",
       xlab = "Percent", # becomes ylab due to horizontal arg,
       ylab = c("Before","After"),
       title = "Overall, this was an excellent course.",
       auto.key = list(space = "right", columns = 1,
                     reverse = TRUE))
```

## Graded assignments reflected the material covered

```{r reflected-material}
overall.data <-
  te.data %>%
  filter(Number==893) %>%
  mutate(Item=as.factor(Year))

plot.data <- overall.data[7:3]
rownames(plot.data) <- c("Before","After")

likert(plot.data, horizontal = FALSE,
       main = "Graded assignments reflected the material covered.", # or give "title",
       xlab = "Percent", # becomes ylab due to horizontal arg,
       ylab = c("Before","After"),
       title = "Overall, this was an excellent course.",
       auto.key = list(space = "right", columns = 1,
                     reverse = TRUE))
```

## Knew what was expected of me

```{r knew-expectations}
overall.data <-
  te.data %>%
  filter(Number==1633) %>%
  mutate(Item=as.factor(Year))

plot.data <- overall.data[7:3]
rownames(plot.data) <- c("Before","After")

likert(plot.data, horizontal = FALSE,
       main = "I knew what was expected of me in this course..", # or give "title",
       xlab = "Percent", # becomes ylab due to horizontal arg,
       ylab = c("Before","After"),
       title = "Overall, this was an excellent course.",
       auto.key = list(space = "right", columns = 1,
                     reverse = TRUE))
```


# Workload



```{r workload}
overall.data <-
  te.data %>%
  filter(Number==891) %>%
  mutate(Item=as.factor(Year))

plot.data <- overall.data[7:3]
rownames(plot.data) <- c("Before","After")

likert(plot.data, horizontal = FALSE,
       main = "As compared with other courses of equal credit, the workload of this course was:", # or give "title",
       xlab = "Percent", # becomes ylab due to horizontal arg,
       ylab = c("Before","After"),
       title = "Overall, this was an excellent course.",
       auto.key = list(space = "right", columns = 1,
                     reverse = TRUE))
```