\documentclass{tufte-handout}

%\geometry{showframe}% for debugging purposes -- displays the margins

\usepackage{amsmath}

% Set up the images/graphics package
\usepackage{graphicx}
\setkeys{Gin}{width=\linewidth,totalheight=\textheight,keepaspectratio}
\graphicspath{{graphics/}}

\title{Endocrine and Hypothalamic Control of Appetite}
\author{Dave Bridges, Ph.D.}
%\date{24 January 2009}  % if the \date{} command is left out, the current date will be used

% The following package makes prettier tables.  We're all about the bling!
\usepackage{booktabs}

% The units package provides nice, non-stacked fractions and better spacing
% for units.
\usepackage{units}

% The fancyvrb package lets us customize the formatting of verbatim
% environments.  We use a slightly smaller font.
\usepackage{fancyvrb}
\fvset{fontsize=\normalsize}

% Small sections of multiple columns
\usepackage{multicol}

% Provides paragraphs of dummy text
\usepackage{lipsum}

% These commands are used to pretty-print LaTeX commands
\newcommand{\doccmd}[1]{\texttt{\textbackslash#1}}% command name -- adds backslash automatically
\newcommand{\docopt}[1]{\ensuremath{\langle}\textrm{\textit{#1}}\ensuremath{\rangle}}% optional command argument
\newcommand{\docarg}[1]{\textrm{\textit{#1}}}% (required) command argument
\newenvironment{docspec}{\begin{quote}\noindent}{\end{quote}}% command specification environment
\newcommand{\docenv}[1]{\textsf{#1}}% environment name
\newcommand{\docpkg}[1]{\texttt{#1}}% package name
\newcommand{\doccls}[1]{\texttt{#1}}% document class name
\newcommand{\docclsopt}[1]{\texttt{#1}}% document class option name

\begin{document}

\maketitle% this prints the handout title, author, and date

\begin{abstract}
\noindent This lecture covers endocrine control of appetite.  It covers the following pages in the textbook: XXXX \cite{Widmaier2013}.  T
\end{abstract}

\tableofcontents

\pagebreak

\section{Learning Objectives}
For this lecture, the learning objectives are:
\begin{itemize}
\item Describe the appetite-regulating hormones secreted from the gut, how they are regulated and under what conditions they are released.
\item Describe the AgRP/POMC circuit and its relationship to both circulating factors and neuropeptides.
\item Understand the relationship between adipose mass and appetite regulation, including how adipokines are regulated and what role they play.
\item List the effects of insulin on appetite and what the neurological targets of insulin are.
\item Describe the role of the blood-brain barrier in the regulation of appetite and how it is altered in obesity.
\item Describe how hypothalamic feeding circuits integrate with other pleasure and reward circuits in the brain.
\item Explain how neuroendocrine obesity differs from idiopathic obesity and how they might be treated in different ways.

\end{itemize}



\listoffigures
\listoftables

\bibliography{library}
\bibliographystyle{plainnat}



\end{document}
