---
title: "Scoring Evaluations"
author: "Dave Bridges"
date: "1/31/2018"
output:
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)

opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```

## Data Import

Exported **Assignment Type Summaries** from GradeCraft and imported.

```{r data-import}
library(readr)
library(dplyr)
library(tidyr)
datafile <- 'Principles of Nutritional Sciences Assignment Type Summary - 2018-02-01.csv'

library(readr)
dataset <- read_csv(datafile)

dropped.students <- c('dave.bridges','ajian','zhongyli')

assessment.dataset <-
  dataset %>%
  select(-`First Name`, -`Last Name`, -Email, -Team) %>%
  gather(Assignment, Points, -Username) %>% 
  filter(!(Username %in% dropped.students)) %>%
  arrange(Points)

summary.dataset <-
  assessment.dataset %>%
  group_by(Username) %>%
  summarize(Total = sum(Points)) %>%
  arrange(-Total)
```

This script imports the data from **`r datafile`**.  This script is located at `r getwd()` and was most recently run on `r date()`

```{r points-distribution}
library(ggplot2)

p <- ggplot(data=summary.dataset, aes(x=Total)) 
p + geom_histogram(binwidth=50) +
  labs(x="Final Points",y="Number of Students", title="Distribution of Final Points") +
  geom_vline(aes(xintercept=1000), color="red", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=950), color="red", linetype="dashed", size=0.5) +
  geom_vline(aes(xintercept=900), color="red", linetype="dashed", size=0.5) +
  geom_vline(aes(xintercept=850), color="red", linetype="dashed", size=0.5) +
  geom_vline(aes(xintercept=800), color="red", linetype="dashed", size=0.5) +
  geom_vline(aes(xintercept=750), color="red", linetype="dashed", size=0.5) +
  geom_vline(aes(xintercept=700), color="red", linetype="dashed", size=0.5) 
```

There were `r filter(summary.dataset, Total>1000) %>% count` students who earned an A.  Among those who earned an A, their average points were `r  filter(summary.dataset, Total>1000) %>% summarize(Averagel=mean(Total))`.  This means that the A students averaged **`r  filter(summary.dataset, Total>1000) %>% summarize(Averagel=mean(Total))-1000`** points too many than they needed.  This is relative to the overall mean of `r summary.dataset %>% summarize(Averagel=mean(Total))`

## Assessment Choices

### Aggregated

```{r points-from-assessments-summary}
required.assignments <- c('Midterm','Class Based Assignments','In Class Quiz')

assignment.summary.data <-
  assessment.dataset %>%
  group_by(Assignment) %>%
  summarize(Mean.Points = mean(Points),
            SD.Points = sd(Points)) %>%
  mutate(Required = ifelse(Assignment %in% required.assignments, "Required", "Optional")) %>%
  mutate(Required = relevel(as.factor(Required), ref="Required")) %>%
  arrange(Required, -Mean.Points)

assignment.summary.data.required <-
  assessment.dataset %>%
  mutate(Required = ifelse(Assignment %in% required.assignments, "Required", "Optional")) %>%
  mutate(Required = relevel(as.factor(Required), ref="Required")) %>%
  group_by(Username, Required) %>%
  summarize(Points = sum(Points)) %>%
  group_by(Required) %>%
  summarize(Median.Points = median(Points),
            Median.Points.SD = sd(Points)) %>%
  mutate(Relative.Points = Median.Points/sum(Median.Points) *100,
         Relative.Points.SD = Median.Points.SD/sum(Median.Points)*100)

assignment.summary.data.relative <-
  assessment.dataset %>%
  group_by(Username) %>%
  mutate(Total = sum(Points)) %>%
  group_by(Assignment) %>%
  summarize(Relative.Points = mean(Points/Total)*100,
            SD.Relative.Points = sd(Points/Total)*100) %>%
  mutate(Required = ifelse(Assignment %in% required.assignments, "Required", "Optional")) %>%
  mutate(Required = relevel(as.factor(Required), ref="Required")) %>%
  arrange(Required, -Relative.Points)

positions <- as.character(assignment.summary.data$Assignment)

library(forcats)

assignment.summary.data$Assignment <- fct_relevel(assignment.summary.data$Assignment, positions)
assignment.summary.data.relative$Assignment <- fct_relevel(assignment.summary.data.relative$Assignment, positions)

p <- ggplot(assignment.summary.data, aes(y=Mean.Points, x=Assignment))
p + geom_bar(stat='identity', aes(fill=Assignment,col=Required)) + scale_x_discrete(limits = positions) +   
  labs(title="Aggregate Assignment Breakdown",  y = "Average Points") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p <- ggplot(filter(assignment.summary.data, Assignment != 'Midterm'), aes(y=Mean.Points, x=Assignment))
p + geom_bar(stat='identity', aes(fill=Assignment,col=Required)) + scale_x_discrete(limits = positions) +   
  labs(title="Aggregate Assignment Breakdown - Excluding Midterm", y = "Average Points") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p <- ggplot(assignment.summary.data.relative, aes(y=Relative.Points, x=Assignment))
p + geom_bar(stat='identity', aes(fill=Assignment,col=Required)) + scale_x_discrete(limits = positions) +   
  labs(title="Aggregate Assignment Breakdown",  y = "Points as a Percent of Total") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p <- ggplot(filter(assignment.summary.data.relative, Assignment != 'Midterm'), aes(y=Relative.Points, x=Assignment))
p + geom_bar(stat='identity', aes(fill=Assignment,col=Required)) + scale_x_discrete(limits = positions) +   
  labs(title="Aggregate Assignment Breakdown - Excluding Midterm", y = "Points as a Percent of Total") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

kable(assignment.summary.data.required, caption="Average total points")
kable(assignment.summary.data, caption="Average total points per assignment")
kable(assignment.summary.data.relative, caption="Average Relative points per assignment as a percent of total")

```

### Student Level

```{r points-from-assessments-individual}
assessment.dataset$Username <- fct_relevel(assessment.dataset$Username, summary.dataset$Username)  
assessment.dataset$Assignment <- fct_relevel(assessment.dataset$Assignment, positions)

p <- ggplot(assessment.dataset, aes(y=Points,x=Username))

p + geom_bar(stat='identity', aes(fill=Assignment)) +
  labs(title="Student Level Assignment Breakdown") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
    geom_hline(aes(yintercept=1000), color="black", linetype="dashed", size=0.5) 
```

#### Student Level Paths

```{r correlations-all}
assignment.matrix <- assessment.dataset %>% spread(Assignment,Points)
assignment.matrix.names <- assignment.matrix$Username

assignment.matrix <- 
  assignment.matrix %>% 
  select(-Username) %>% 
  as.matrix()

library(RColorBrewer)
library(gplots)
heatmap.2(assignment.matrix, 
        scale='col',
        labRow=NA,
        cexCol = 0.3,
        col=brewer.pal(9,"Blues"),
        density.info='none', trace='none')

heatmap.2(assignment.matrix, 
        scale='col',
        labRow=NA,
        Colv=NA,
        cexCol = 0.3,
        col=brewer.pal(9,"Blues"),
        density.info='none', trace='none')
```

# Among Students with an A

```{r correlations-a-only}
assignment.matrix.a <- assessment.dataset %>% spread(Assignment,Points)
assignment.matrix.a$Total <- rowSums(assignment.matrix.a %>% select(-Username))

assignment.matrix.a <- 
  assignment.matrix.a %>% 
  filter(Total>1000) %>%
  select(-Username, -Total) %>% 
  as.matrix()

heatmap.2(assignment.matrix.a, 
        scale='col',
        labRow=NA,
        cexCol = 0.3,
        col=brewer.pal(9,"Blues"),
        density.info='none', trace='none')

heatmap.2(assignment.matrix.a, 
        scale='col',
        labRow=NA,
        Colv=NA,
        cexCol = 0.3,
        col=brewer.pal(9,"Blues"),
        density.info='none', trace='none')
```

```{r required-optional-correlations}
assessment.dataset.required <-
  assessment.dataset %>% 
  mutate(Required = ifelse(Assignment %in% required.assignments, "Required", "Optional")) %>%
  group_by(Username, Required) %>%
  summarise(Points = sum(Points)) %>%
  spread(Required, Points) %>%
  mutate(Total.Points = Required + Optional) %>%
  mutate(A.Grade = Total.Points>1000)
  
lm.points <- lm(Optional ~ Required, data=assessment.dataset.required)  
par(mfrow=c(2,2))
plot(lm.points)

library(broom)
kable(tidy(lm.points), caption="Linear model of optional points dependent on required points per student.")

shapiro.tidy <- function(x) tidy(shapiro.test(x))

kable(with(assessment.dataset.required, rbind(Optional = shapiro.tidy(Optional), 
                                              Required=shapiro.tidy(Required),
                                              Total.Points = shapiro.tidy(Total.Points))), caption="Shapiro-Wilk tests for normality")

#correlations
kable(tidy(with(assessment.dataset.required, cor.test(Optional, Required,method="spearman")), caption="Correlation between optional and required points.  Spearman is used due to lack of normality of the required points distribution."))

kable(tidy(with(assessment.dataset.required, cor.test(Optional, Total.Points,method="pearson")), caption="Correlation between optional and total points."))

p <- ggplot(assessment.dataset.required, aes(y=Optional,x=Required))
p + geom_point() +
  geom_smooth(method='lm') +
  labs(y="Optional Points", x="Required Points") +
  geom_rug(alpha=0.5)

p <- ggplot(assessment.dataset.required, aes(y=Total.Points,x=Optional))
p + geom_point() +
  geom_smooth(method='lm') +
  labs(y="Total Points", x="Optional Points") +
  geom_hline(yintercept = 1000, lty=2,col="red") +
  geom_rug(alpha=0.5)
```


# Likelyhood of an A Grade

```{r logistic-a, fig.cap="Logistic regression for points required to get an A"}
glm.a.grade <- glm(A.Grade ~ Optional, data=assessment.dataset.required, family="binomial")

tidy(glm.a.grade) %>% kable(caption="Logistic regression of optional assinments vs grade")
#coef as logit

predict.data <- data.frame(Optional=seq(0,400),by=5)
predict.data$Logit <- predict(glm.a.grade, newdata=predict.data)
predict.data <-
  predict.data %>%
  mutate(Odds = exp(Logit)) %>%
  mutate(Probability = Odds/(1+Odds))

ggplot(predict.data, aes(y=Probability,x=Optional)) +
  geom_line(color='cadetblue') +
  labs(y="Probability of getting an A",
       x="Amount of optional points",
       title="Logistic regression of 2018 student data")

beta.a.grade <- coef(glm.a.grade)['Optional'] #coef as logit
odds.a.grade <- exp(beta.a.grade) #coef as odds
prob.a.grade <- odds.a.grade/(1 + odds.a.grade)
```


Based on last year, 

# Session Information

```{r session-info}
sessionInfo()
```