---
title: "Scoring Evaluations"
author: "Dave Bridges"
date: "1/31/2018"
output:
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)

opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```

## Data Import

Exported **Assignment Type Summaries** from GradeCraft and imported.

```{r data-import}
library(readr)
library(dplyr)
library(tidyr)
datafile <- 'Principles of Nutritional Sciences Assignment Type Summary - 2018-02-01.csv'

library(readr)
dataset <- read_csv(datafile)

dropped.students <- c('dave.bridges','ajian','zhongyli')

assessment.dataset <-
  dataset %>%
  select(-`First Name`, -`Last Name`, -Email, -Team) %>%
  gather(Assignment, Points, -Username) %>% 
  filter(!(Username %in% dropped.students)) %>%
  arrange(Points)

summary.dataset <-
  assessment.dataset %>%
  group_by(Username) %>%
  summarize(Total = sum(Points)) %>%
  arrange(-Total)
```

This script imports the data from **`r datafile`**.  This script is located at `r getwd()` and was most recently run on `r date()`

```{r points-distribution}
library(ggplot2)

p <- ggplot(data=summary.dataset, aes(x=Total)) 
p + geom_histogram(binwidth=50) +
  labs(x="Final Points",y="Number of Students", title="Distribution of Final Points") +
  geom_vline(aes(xintercept=1000), color="red", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=950), color="red", linetype="dashed", size=0.5) +
  geom_vline(aes(xintercept=900), color="red", linetype="dashed", size=0.5) +
  geom_vline(aes(xintercept=850), color="red", linetype="dashed", size=0.5) +
  geom_vline(aes(xintercept=800), color="red", linetype="dashed", size=0.5) +
  geom_vline(aes(xintercept=750), color="red", linetype="dashed", size=0.5) +
  geom_vline(aes(xintercept=700), color="red", linetype="dashed", size=0.5) 
```

There were `r filter(summary.dataset, Total>1000) %>% count` students who earned an A.  Among those who earned an A, their average points were `r  filter(summary.dataset, Total>1000) %>% summarize(Averagel=mean(Total))`.  This means that the A students averaged **`r  filter(summary.dataset, Total>1000) %>% summarize(Averagel=mean(Total))-1000`** points too many than they needed.  This is relative to the overall mean of `r summary.dataset %>% summarize(Averagel=mean(Total))`

## Assessment Choices

### Aggregated

```{r points-from-assessments-summary}
required.assignments <- c('Midterm','Class Based Assignments','In Class Quiz')

assignment.summary.data <-
  assessment.dataset %>%
  group_by(Assignment) %>%
  summarize(Mean.Points = mean(Points),
            SD.Points = sd(Points)) %>%
  mutate(Required = ifelse(Assignment %in% required.assignments, "Required", "Optional")) %>%
  mutate(Required = relevel(as.factor(Required), ref="Required")) %>%
  arrange(Required, -Mean.Points)

assignment.summary.data.required <-
  assessment.dataset %>%
  mutate(Required = ifelse(Assignment %in% required.assignments, "Required", "Optional")) %>%
  mutate(Required = relevel(as.factor(Required), ref="Required")) %>%
  group_by(Username, Required) %>%
  summarize(Points = sum(Points)) %>%
  group_by(Required) %>%
  summarize(Median.Points = median(Points),
            Median.Points.SD = sd(Points)) %>%
  mutate(Relative.Points = Median.Points/sum(Median.Points) *100,
         Relative.Points.SD = Median.Points.SD/sum(Median.Points)*100)

assignment.summary.data.relative <-
  assessment.dataset %>%
  group_by(Username) %>%
  mutate(Total = sum(Points)) %>%
  group_by(Assignment) %>%
  summarize(Relative.Points = mean(Points/Total)*100,
            SD.Relative.Points = sd(Points/Total)*100) %>%
  mutate(Required = ifelse(Assignment %in% required.assignments, "Required", "Optional")) %>%
  mutate(Required = relevel(as.factor(Required), ref="Required")) %>%
  arrange(Required, -Relative.Points)

positions <- as.character(assignment.summary.data$Assignment)

library(forcats)

assignment.summary.data$Assignment <- fct_relevel(assignment.summary.data$Assignment, positions)
assignment.summary.data.relative$Assignment <- fct_relevel(assignment.summary.data.relative$Assignment, positions)

p <- ggplot(assignment.summary.data, aes(y=Mean.Points, x=Assignment))
p + geom_bar(stat='identity', aes(fill=Assignment,col=Required)) + scale_x_discrete(limits = positions) +   
  labs(title="Aggregate Assignment Breakdown",  y = "Average Points") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p <- ggplot(filter(assignment.summary.data, Assignment != 'Midterm'), aes(y=Mean.Points, x=Assignment))
p + geom_bar(stat='identity', aes(fill=Assignment,col=Required)) + scale_x_discrete(limits = positions) +   
  labs(title="Aggregate Assignment Breakdown - Excluding Midterm", y = "Average Points") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p <- ggplot(assignment.summary.data.relative, aes(y=Relative.Points, x=Assignment))
p + geom_bar(stat='identity', aes(fill=Assignment,col=Required)) + scale_x_discrete(limits = positions) +   
  labs(title="Aggregate Assignment Breakdown",  y = "Points as a Percent of Total") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p <- ggplot(filter(assignment.summary.data.relative, Assignment != 'Midterm'), aes(y=Relative.Points, x=Assignment))
p + geom_bar(stat='identity', aes(fill=Assignment,col=Required)) + scale_x_discrete(limits = positions) +   
  labs(title="Aggregate Assignment Breakdown - Excluding Midterm", y = "Points as a Percent of Total") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

kable(assignment.summary.data.required, caption="Average total points")
kable(assignment.summary.data, caption="Average total points per assignment")
kable(assignment.summary.data.relative, caption="Average Relative points per assignment as a percent of total")

```

### Student Level

```{r points-from-assessments-individual}
assessment.dataset$Username <- fct_relevel(assessment.dataset$Username, summary.dataset$Username)  
assessment.dataset$Assignment <- fct_relevel(assessment.dataset$Assignment, positions)

p <- ggplot(assessment.dataset, aes(y=Points,x=Username))

p + geom_bar(stat='identity', aes(fill=Assignment)) +
  labs(title="Student Level Assignment Breakdown") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
    geom_hline(aes(yintercept=1000), color="black", linetype="dashed", size=0.5) 
```
