---
title: "Summary Graphs of NUTR630 Post-Semester Feedback"
author: "Dave Bridges, Liv Anderson and Rina Hisamatsu"
date: "2017-09-03"
output:
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: no
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```


```{r data-input}
library(readr)
onboarding.filename <- 'https://docs.google.com/spreadsheets/d/e/2PACX-1vQI-b1A4Zd-gx3FuY3lkKbWE0zKUfmAhpFovKOxow5AC1wWQpBsSvOKI0_gtZ4DJu5sj-YM1_1nsKUe/pub?gid=1032535237&single=true&output=csv'
onboarding.data <- read_csv(onboarding.filename)

end.filename <- 'https://docs.google.com/spreadsheets/d/e/2PACX-1vTxPKhKOwkZ18U1X0Z44M8wGqtM5ow18tVPF4yupZbzditOQdlQ3DJkfJJA-Dl226NfGYJ0tEUqOFn4/pub?gid=1563141647&single=true&output=csv'
end.data <- read_csv(end.filename)

merged.data <- full_join(onboarding.data,end.data, by='Email Address', suffix=c("before","after"))
```

These data can be found in `r getwd()`.  This script was most recently updated on `r date()`.

# Analysis

## Interests

## Learning Activities

### Thoughts about whether they learned a lot from the activity

```{r learning-activities}
library(dplyr)

mylevels <- c('Strongly Agree','Agree','Neutral','Disagree','Strongly Disagree')


activity.data <- 
  end.data %>%
  dplyr::select(contains("learned")) %>%
  rename(`Written Reports` = `Select one in reference to written reports [I learned a lot from this activity]`,
         `Presentation` = `Select one in reference to your presentation [I learned a lot from this activity]`,
         `Peer Reviews` = `Select one answer in reference to the peer reviews [I learned a lot from this activity]`,
         `Social Media` = `Select one in reference to the social media assignments [I learned a lot from this activity]`,
         `News Posts` = `Select one in reference to the news posts [I learned a lot from this activity]`,
         `Review Questions` = `Select one in reference to the review questions [I learned a lot from this activity]`,
         `Problem Roulette` = `Select one answer in reference to problem roulette [I learned a lot from this activity]`) %>%
  mutate(`Written Reports` = factor(`Written Reports`,levels=mylevels),
         `Presentation`= factor(`Presentation`, levels=mylevels),
         `Peer Reviews`= factor(`Peer Reviews`, levels=mylevels),
         `Social Media`= factor(`Social Media`, levels=mylevels),
         `News Posts`= factor(`News Posts`, levels=mylevels),
         `Review Questions`= factor(`Review Questions`, levels=mylevels),
         `Problem Roulette`= factor(`Problem Roulette`, levels=mylevels))

library(sjPlot)
sjp.likert(activity.data, title="I learned a lot from this activity")
```

### Thoughts about whether these assesed knowledge

```{r assessed-knowledge}
assessed.data <- 
  end.data %>%
  dplyr::select(contains("assessed")) %>%
  rename(`Written Reports` = `Select one in reference to written reports [This effectively assessed my knowledge]`,
         `Presentation` = `Select one in reference to your presentation [This effectively assessed my knowledge]`,
         `Peer Reviews` = `Select one answer in reference to the peer reviews [This effectively assessed my knowledge]`,
         `Social Media` = `Select one in reference to the social media assignments [This effectively assessed my knowledge]`,
         `News Posts` = `Select one in reference to the news posts [This effectively assessed my knowledge]`,
         `Review Questions` = `Select one in reference to the review questions [This effectively assessed my knowledge]`,
         `Problem Roulette` = `Select one answer in reference to problem roulette [This effectively assessed my knowledge]`) %>%
  mutate(`Written Reports` = factor(`Written Reports`,levels=mylevels),
         `Presentation`= factor(`Presentation`, levels=mylevels),
         `Peer Reviews`= factor(`Peer Reviews`, levels=mylevels),
         `Social Media`= factor(`Social Media`, levels=mylevels),
         `News Posts`= factor(`News Posts`, levels=mylevels),
         `Review Questions`= factor(`Review Questions`, levels=mylevels),
         `Problem Roulette`= factor(`Problem Roulette`, levels=mylevels))

sjp.likert(assessed.data, title="This effectively assessed my knowledge")
```

### Thoughts as to whether it was worth an ppropriate amount of points

```{r amount-of-points}
points.data <- 
  end.data %>%
  dplyr::select(contains("appropriate amount")) %>%
  rename(`Written Reports` = `Select one in reference to written reports [This was worth an appropriate amount of points based on the amount of effort required]`,
         `Presentation` = `Select one in reference to your presentation [This was worth an appropriate amount of points based on the amount of effort required]`,
         `Peer Reviews` = `Select one answer in reference to the peer reviews [This was worth an appropriate amount of points based on the amount of effort required]`,
         `Social Media` = `Select one in reference to the social media assignments [This was worth an appropriate amount of points based on the amount of effort required]`,
         `News Posts` = `Select one in reference to the news posts [This was worth an appropriate amount of points based on the amount of effort required]`,
         `Review Questions` = `Select one in reference to the review questions [This was worth an appropriate amount of points based on the amount of effort required]`,
         `Problem Roulette` = `Select one answer in reference to problem roulette [This was worth an appropriate amount of points based on the amount of effort required]`) %>%
  mutate(`Written Reports` = factor(`Written Reports`,levels=mylevels),
         `Presentation`= factor(`Presentation`, levels=mylevels),
         `Peer Reviews`= factor(`Peer Reviews`, levels=mylevels),
         `Social Media`= factor(`Social Media`, levels=mylevels),
         `News Posts`= factor(`News Posts`, levels=mylevels),
         `Review Questions`= factor(`Review Questions`, levels=mylevels),
         `Problem Roulette`= factor(`Problem Roulette`, levels=mylevels))

sjp.likert(points.data, title="This was worth an appropriate amount of points based on the amount of effort required")
```

# Changes over the semester

## Interests and Aptitudes

```{r interests}
merged.data[merged.data==1]<-'Strongly Agree'
merged.data[merged.data==2]<-'Agree'
merged.data[merged.data==3]<-'Neutral'
merged.data[merged.data==4]<-'Disagree'
merged.data[merged.data==5]<-'Strongly Disagree'


macronutritent.aptitude.data <- 
  merged.data %>%
  dplyr::select(contains("macronutrient")) %>%
  rename(`Interested Before` = `Macronutrient biochemistry is of interest to me.before`,
         `Interested After` = `Macronutrient biochemistry is of interest to me.after`,
         `Important Before` = `Macronutrient biochemistry is important for my career interests.before`,
         `Important After` = `Macronutrient biochemistry is important for my career interests.after`) %>%
  mutate(`Interested Before`= factor(`Interested Before`, levels=mylevels),
         `Interested After`= factor(`Interested After`, levels=mylevels),
         `Important Before`= factor(`Important Before`, levels=mylevels),
         `Important After`= factor(`Important After`, levels=mylevels))
col.order <- c(1,3,2,4)

sjp.likert(macronutritent.aptitude.data[,col.order], title="Macronutrient biochemistry")


digestive.aptitude.data <- 
  merged.data %>%
  dplyr::select(contains("digestive")) %>%
  rename(`Interested Before` = `Comprehensive understanding of the digestive tract is of interest to me.before`,
         `Interested After` = `Comprehensive understanding of the digestive tract is of interest to me.after`,
         `Important Before` = `Comprehensive understanding of the digestive tract is important for my career interests.before`,
         `Important After` = `Comprehensive understanding of the digestive tract is important for my career interests.after`) %>%
  mutate(`Interested Before`= factor(`Interested Before`, levels=mylevels),
         `Interested After`= factor(`Interested After`, levels=mylevels),
         `Important Before`= factor(`Important Before`, levels=mylevels),
         `Important After`= factor(`Important After`, levels=mylevels))
col.order <- c(1,3,2,4)

sjp.likert(digestive.aptitude.data[,col.order], title="Understanding of the digestive tract")
```

# Metacognition

```{r metacognition}
meta.data <- 
  merged.data %>%
  dplyr::select(contains("What is your level of agreement on the following statements? [When given an assignment to complete, I produce better outcomes with a community of support (i.e., instructor or peer support).]")) %>%
  rename(`Before` = `What is your level of agreement on the following statements? [When given an assignment to complete, I produce better outcomes with a community of support (i.e., instructor or peer support).]before`,
         `After` =  `What is your level of agreement on the following statements? [When given an assignment to complete, I produce better outcomes with a community of support (i.e., instructor or peer support).]after`) %>%
  mutate(`Before`= factor(`Before`, levels=mylevels),
         `After`= factor(`After`, levels=mylevels))

sjp.likert(meta.data, title="When given an assignment to complete, I produce better outcomes with a community of support (i.e., instructor or peer support)", cat.neutral='Neutral')


meta.data <- 
  merged.data %>%
  dplyr::select(contains("What is your level of agreement on the following statements? [I am motivated to complete an assignment if I enjoy doing it.]")) %>%
  rename(`Before` = `What is your level of agreement on the following statements? [I am motivated to complete an assignment if I enjoy doing it.]before`,
         `After` =  `What is your level of agreement on the following statements? [I am motivated to complete an assignment if I enjoy doing it.]after`) %>%
  mutate(`Before`= factor(`Before`, levels=mylevels),
         `After`= factor(`After`, levels=mylevels))

sjp.likert(meta.data, title="I am motivated to complete an assignment if I enjoy doing it.", cat.neutral='Neutral')


meta.data <- 
  merged.data %>%
  dplyr::select(contains("What is your level of agreement on the following statements? [I am aware of the skills I develop when I complete different types of assignments.]")) %>%
  rename(`Before` = `What is your level of agreement on the following statements? [I am aware of the skills I develop when I complete different types of assignments.]before`,
         `After` =  `What is your level of agreement on the following statements? [I am aware of the skills I develop when I complete different types of assignments.]after`) %>%
  mutate(`Before`= factor(`Before`, levels=mylevels),
         `After`= factor(`After`, levels=mylevels))

sjp.likert(meta.data, title="I am aware of the skills I develop when I complete different types of assignments.", cat.neutral='Neutral')


meta.data <- 
  merged.data %>%
  dplyr::select(contains("What is your level of agreement on the following statements? [When a difficult assignment is given to me, I feel confident that I can complete it.]")) %>%
  rename(`Before` = `What is your level of agreement on the following statements? [When a difficult assignment is given to me, I feel confident that I can complete it.]before`,
         `After` =  `What is your level of agreement on the following statements? [When a difficult assignment is given to me, I feel confident that I can complete it.]after`) %>%
  mutate(`Before`= factor(`Before`, levels=mylevels),
         `After`= factor(`After`, levels=mylevels))

sjp.likert(meta.data, title="When a difficult assignment is given to me, I feel confident that I can complete it", cat.neutral='Neutral')


meta.data <- 
  merged.data %>%
  dplyr::select(contains("What is your level of agreement on the following statements? [I perform better when completing an assignment that I enjoy (i.e., get a better grade).]")) %>%
  rename(`Before` = `What is your level of agreement on the following statements? [I perform better when completing an assignment that I enjoy (i.e., get a better grade).]before`,
         `After` =  `What is your level of agreement on the following statements? [I perform better when completing an assignment that I enjoy (i.e., get a better grade).]after`) %>%
  mutate(`Before`= factor(`Before`, levels=mylevels),
         `After`= factor(`After`, levels=mylevels))

sjp.likert(meta.data, title="I perform better when completing an assignment that I enjoy (i.e., get a better grade).", cat.neutral='Neutral')

```

# Session Information

```{r session-information, echo=T}
sessionInfo()
```